---
title: "Cell Distance Analysis — Spatial Neighbourhood Enrichment (Δ%)"
author: "Victoire Ringler"
date: "2026-01-09"
output: html_document
---

## Overview

This pipeline quantifies the **spatial neighbourhood composition** of malignant
B cells (stratified by M26 marker status) relative to other cell types in
multiplexed imaging data. For each tissue core, it:

1. Loads spatial point pattern data (`.rds` files) and annotates cells with
   their type and M26 status
2. Computes the **k-nearest neighbour (k=50) cell type proportions** for every
   cell in each core
3. Calculates **Δ%** — the difference between local neighbourhood proportions
   and global core proportions — as a measure of spatial enrichment/depletion
4. Compares neighbourhood composition between:
   - **Primary vs Relapse** biopsies
   - **Dispersed vs Clustered** tumour cores (tertile-stratified by clustering
     index, Primary only)
5. Exports summary statistics and publication-quality histograms as PDFs

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# ============================================================
# DEPENDENCIES
# ============================================================
# BiocManager::install("zellkonverter")  # Uncomment to install
# install.packages("readxl")             # Uncomment to install

library(reticulate)
use_condaenv("zellenv", required = TRUE)  # Python environment for zellkonverter

library(readxl)           # Read Excel metadata files
library(zellkonverter)    # Convert between AnnData and SingleCellExperiment
library(spatstat.geom)    # Spatial point pattern geometry
library(spatstat.explore) # Spatial statistics and nearest-neighbour functions
library(dplyr)            # Data manipulation
library(tidyr)            # Data reshaping
library(stringr)          # String operations for parsing file/condition names
library(purrr)            # Functional programming (map, lapply wrappers)
library(ggplot2)          # Plotting
```

---

## 2. Load Cell Metadata

```{r load-cell-data}
# ============================================================
# LOAD CELL METADATA FROM ANNDATA OBS CSV
# ============================================================
# This CSV is exported from the AnnData object's .obs dataframe and contains
# spatial coordinates, cell type annotations, and M26 marker status per cell.

cell_df <- read.csv("/Users/victoireringler/Documents/Singapore/Thesis/adata_obs.csv")

# Keep only the columns needed for spatial analysis
cell_df <- cell_df[, c("x", "y", "condition", "cell_type_annotation", "M26")]

# Parse Image_ID from the condition string by stripping the trailing
# coordinate bracket suffix, e.g.:
#   "CT20-09 A1_040625_Core[1,1,A]_[7546,39932]" -> "CT20-09 A1_040625_Core[1,1,A]"
# This allows matching cells to their corresponding RDS point pattern files.
cell_df <- cell_df %>%
  mutate(
    Image_ID = str_replace(
      as.character(condition),
      "_\\[.*\\]$",   # Remove trailing _[x,y] coordinate suffix
      ""
    )
  ) %>%
  select(x, y, Image_ID, cell_type_annotation, M26)

head(cell_df)
```

---

## 3. Load Spatial Point Patterns & Annotate Marks

```{r load-ppp-files}
# ============================================================
# HELPER: REPLACE PPP MARKS WITH CELL TYPE + M26 STATUS
# ============================================================
# Each RDS file is a spatstat ppp object whose marks need to be replaced
# with a combined label:
#   - Malignant B cells with M26 == "M+2+6-"  -> "Malignant_B_M26"
#   - All other Malignant B cells              -> "Malignant_B_other"
#   - All other cell types                     -> their original annotation

replace_ppp_marks <- function(ppp_obj, cell_df, Image_ID) {
  # Subset cell metadata to the current core
  df_core <- cell_df %>% filter(Image_ID == Image_ID)

  # Extract spatial coordinates from the ppp object for matching
  ppp_df <- data.frame(x = ppp_obj$x, y = ppp_obj$y)

  # Join cell metadata by exact x,y coordinate match
  joined <- left_join(ppp_df, df_core, by = c("x", "y"))

  # Create combined mark label
  joined <- joined %>%
    mutate(
      new_mark = case_when(
        cell_type_annotation == "Malignant B cell" & M26 == "M+2+6-" ~ "Malignant_B_M26",
        cell_type_annotation == "Malignant B cell"                    ~ "Malignant_B_other",
        TRUE                                                           ~ cell_type_annotation
      )
    )

  marks(ppp_obj) <- factor(joined$new_mark)
  ppp_obj
}

# ============================================================
# LOAD ALL RDS FILES AND ANNOTATE
# ============================================================
ppp_files <- list.files(
  "/Users/victoireringler/Documents/Singapore/Thesis/BCC_PatternExtract/M26/final/",
  pattern    = "\\.[rR][dD][sS]$",
  full.names = TRUE
)

ppp_list <- lapply(ppp_files, function(f) {
  # Parse Image_ID from filename (same stripping logic as for cell_df)
  condition <- tools::file_path_sans_ext(basename(f))
  Image_ID  <- str_replace(condition, "_\\[.*\\]$", "")

  ppp <- readRDS(f)
  replace_ppp_marks(ppp_obj = ppp, cell_df = cell_df, Image_ID = Image_ID)
})

# Name list elements by Image_ID for easy lookup downstream
names(ppp_list) <- sapply(ppp_files, function(f) {
  condition <- tools::file_path_sans_ext(basename(f))
  str_replace(condition, "_\\[.*\\]$", "")
})
```

---

## 4. Compute k-Nearest Neighbour Proportions

```{r calculate-proportions}
# ============================================================
# KNN NEIGHBOURHOOD COMPOSITION
# ============================================================
k <- 50  # Number of nearest neighbours to consider per cell

# Collect all unique cell types across all cores for a consistent column set
unique_types_all <- unique(unlist(lapply(ppp_list, function(ppp) marks(ppp))))

# Remove cells with NA marks (can occur if coordinate matching failed)
sanitize_ppp <- function(ppp_obj) {
  keep <- !is.na(marks(ppp_obj))
  ppp_obj[keep]
}
ppp_list <- map(ppp_list, sanitize_ppp)

# For each core, compute the proportion of each cell type
# among the k nearest neighbours of every cell.
proportions_list <- lapply(ppp_list, function(ppp_obj) {

  # nn_idx: matrix of shape (n_cells × k), each row = indices of k nearest neighbours
  nn_idx <- nnwhich(ppp_obj, k = 1:k)

  # nn_types: matrix of shape (n_cells × k), each entry = cell type of that neighbour
  nn_types <- apply(nn_idx, 2, function(col) as.character(marks(ppp_obj)[col]))

  # prop_mat: matrix of shape (n_cells × n_cell_types)
  # Each row = proportion of each cell type among the k neighbours of that cell
  prop_mat <- t(apply(nn_types, 1, function(neighbors) {
    sapply(unique_types_all, function(t) mean(neighbors == t))
  }))

  colnames(prop_mat) <- unique_types_all
  prop_mat
})
```

---

## 5. Calculate Mean Proportions and Δ% per Core

```{r calculate-delta}
# ============================================================
# DELTA % = LOCAL NEIGHBOURHOOD PROPORTION - GLOBAL CORE PROPORTION
# ============================================================
# For each focal cell type in a core:
#   - mean_neighbor_proportion: average proportion of each neighbour cell type
#     among the k nearest neighbours of all focal cells
#   - overall_proportion: global frequency of that neighbour cell type in the core
#   - delta_percent: difference (positive = locally enriched, negative = depleted)
#   - delta_percent_abs: absolute difference (magnitude of enrichment regardless of direction)

calculate_mean_proportions <- function(ppp_obj, prop_mat) {
  cell_types  <- as.character(marks(ppp_obj))
  focal_types <- unique(cell_types)

  # Global cell type frequencies in this core
  overall_props_table <- table(cell_types) / length(cell_types)

  # Ensure all possible cell types are represented (fill missing with 0)
  all_cell_types <- colnames(prop_mat)
  overall_props  <- setNames(rep(0, length(all_cell_types)), all_cell_types)
  overall_props[names(overall_props_table)] <- as.numeric(overall_props_table)

  # Compute per-focal-type summary
  results <- lapply(focal_types, function(focal_type) {
    focal_idx <- which(cell_types == focal_type)
    if (length(focal_idx) == 0) return(NULL)

    # Average neighbour composition for cells of this focal type
    neighbor_props <- colMeans(prop_mat[focal_idx, , drop = FALSE])

    # Retrieve matching global proportions
    overall_for_neighbors <- overall_props[names(neighbor_props)]

    data.frame(
      focal_cell_type          = focal_type,
      neighbor_cell_type       = names(neighbor_props),
      mean_neighbor_proportion = as.numeric(neighbor_props),
      overall_proportion       = as.numeric(overall_for_neighbors),
      delta_percent_abs        = abs(neighbor_props - overall_for_neighbors),
      delta_percent            = neighbor_props - overall_for_neighbors,
      stringsAsFactors         = FALSE,
      row.names                = NULL
    )
  })

  bind_rows(results)
}

# Run for all cores and combine into a single long-format data frame
mean_proportions_by_core <- lapply(names(ppp_list), function(core_name) {
  result          <- calculate_mean_proportions(ppp_list[[core_name]],
                                                proportions_list[[core_name]])
  result$Image_ID <- core_name
  result
})

all_mean_proportions <- bind_rows(mean_proportions_by_core)
```

---

## 6A. Primary vs Relapse Analysis

### Load & Filter Metadata

```{r load-metadata-occurrence}
# ============================================================
# METADATA: BIOPSY OCCURRENCE (PRIMARY vs RELAPSE)
# ============================================================
metadata <- read_xlsx("/Users/victoireringler/Documents/Singapore/Thesis/BCC_mIHC_analysis/survival_ci_per_core.xlsx")

# Restrict to R-CHOP-based frontline treatment regimens (standard of care for DLBCL)
treatments_kept <- c(
  'CHOP6-RITU', 'CHOP4-RITU', 'CHOP8-RITU', 'CHOP6-RITU-RAD', 'CHOP3-RITU-RAD',
  'CHOP6-RITU-RITUMAIN', 'CHOP-RITU', 'RAD-CHOP6-RITU', 'CHOP1-RITU-DA-EPOCH5-RITU',
  'CHOP6-RITU-HDMETH-SURG', 'CHOP7-RITU', 'CHOP6-RITU-RITUMAIN2', 'CHOP4-RITU-GDP6-RAD',
  'CHOP6-RITU-GDP1-REFU', 'CHOP6-RITU-HDMETH4', 'SURG-CHOP3-RITU', 'CHOP6-RITU-HDMETH3',
  'CHOP6-RITU-HDMETH-RITUMAIN4', 'CHOP6-RITU-RITUMAIN8', 'CHOP4-RITU-ICE2-RITU'
)

metadata_filt <- metadata %>%
  filter(
    Path               == "DLBC",
    Biopsy_Occurrence  %in% c("Primary", "Relapse"),
    `Front line treatment (regimen)` %in% treatments_kept,
    clustering_index   <= 50           # Exclude extreme outlier cores
  )

# Lookup table: Image_ID -> Biopsy_Occurrence + clustering_index
core_to_occurrence <- metadata_filt %>%
  select(Image_ID, Biopsy_Occurrence, clustering_index)

print("Biopsy Occurrence group distribution:")
print(table(metadata_filt$Biopsy_Occurrence))
```

### Merge Metadata

```{r merge-metadata-occurrence}
# Inner join: only keep cores present in both the ppp data and filtered metadata
all_mean_proportions_with_meta <- all_mean_proportions %>%
  inner_join(core_to_occurrence, by = "Image_ID")

cat("\nCores kept:", length(unique(all_mean_proportions_with_meta$Image_ID)),
    "out of", length(unique(all_mean_proportions$Image_ID)), "total cores\n")

excluded_cores <- setdiff(unique(all_mean_proportions$Image_ID),
                          unique(all_mean_proportions_with_meta$Image_ID))
if (length(excluded_cores) > 0) {
  cat("Excluded cores (not in metadata_filt):", length(excluded_cores), "\n")
}
```

### Summary Statistics

```{r summary-occurrence}
# Overall summary (pooled across all cores)
summary_proportions_overall <- all_mean_proportions_with_meta %>%
  group_by(focal_cell_type, neighbor_cell_type) %>%
  summarise(
    mean_neighbor_prop_across_cores = mean(mean_neighbor_proportion, na.rm = TRUE),
    sd_neighbor_prop                = sd(mean_neighbor_proportion,   na.rm = TRUE),
    mean_overall_prop_across_cores  = mean(overall_proportion,       na.rm = TRUE),
    mean_delta_percent_abs          = mean(delta_percent_abs,        na.rm = TRUE),
    mean_delta_percent              = mean(delta_percent,            na.rm = TRUE),
    sd_delta_percent                = sd(delta_percent,              na.rm = TRUE),
    sd_delta_percent_abs            = sd(delta_percent_abs,          na.rm = TRUE),
    n_cores                         = n(),
    .groups = "drop"
  )

# Stratified summary: Primary vs Relapse
summary_proportions_by_occurrence <- all_mean_proportions_with_meta %>%
  filter(Biopsy_Occurrence %in% c("Primary", "Relapse")) %>%
  group_by(focal_cell_type, neighbor_cell_type, Biopsy_Occurrence) %>%
  summarise(
    mean_neighbor_prop   = mean(mean_neighbor_proportion, na.rm = TRUE),
    sd_neighbor_prop     = sd(mean_neighbor_proportion,   na.rm = TRUE),
    mean_overall_prop    = mean(overall_proportion,       na.rm = TRUE),
    mean_delta_percent   = mean(delta_percent,            na.rm = TRUE),
    mean_delta_percent_abs = mean(delta_percent_abs,      na.rm = TRUE),
    sd_delta_percent     = sd(delta_percent,              na.rm = TRUE),
    sd_delta_percent_abs = sd(delta_percent_abs,          na.rm = TRUE),
    n_cores              = n(),
    .groups = "drop"
  )

print(summary_proportions_overall)
print(summary_proportions_by_occurrence)
```

### Histogram Plots — Primary vs Relapse

```{r plot-occurrence}
# ============================================================
# VISUALISE NEIGHBOURHOOD DISTRIBUTIONS: PRIMARY vs RELAPSE
# ============================================================
# Three plots per focal-neighbour pair:
#   p1 — raw mean neighbour proportion
#   p2 — absolute Δ% (magnitude of enrichment)
#   p3 — signed Δ%  (direction: enriched = positive, depleted = negative)

plot_data <- all_mean_proportions_with_meta %>%
  filter(Biopsy_Occurrence %in% c("Primary", "Relapse"))

focal_cell    <- "Malignant_B_M26"  # Focal cell type (adjust as needed)
neighbor_cell <- "Treg"             # Neighbour cell type (adjust as needed)

plot_data_specific <- plot_data %>%
  filter(focal_cell_type  == focal_cell,
         neighbor_cell_type == neighbor_cell)

occurrence_colours      <- c("Primary" = '#6B8E23', "Relapse" = '#D97706')
occurrence_colours_dark <- c("Primary" = '#536F1C', "Relapse" = '#B05F05')

if (nrow(plot_data_specific) > 0) {

  # --- p1: Mean neighbour proportion ---
  p1 <- ggplot(plot_data_specific,
               aes(x = mean_neighbor_proportion, fill = Biopsy_Occurrence,
                   color = Biopsy_Occurrence)) +
    geom_histogram(alpha = 0.4, position = "identity", bins = 50,
                   aes(y = after_stat(density))) +
    geom_density(alpha = 0.2, linewidth = 1.2) +
    scale_fill_manual(values  = occurrence_colours) +
    scale_color_manual(values = occurrence_colours_dark) +
    labs(
      title    = paste("Distribution of", neighbor_cell, "proportions around", focal_cell),
      subtitle = "Primary vs Relapse cores (with KDE)",
      x        = "Mean neighbour proportion",
      y        = "Density",
      fill     = "Biopsy Occurrence",
      color    = "Biopsy Occurrence"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12), legend.position = "top")
  print(p1)

  # --- p2: Absolute Δ% ---
  p2 <- ggplot(plot_data_specific,
               aes(x = delta_percent_abs, fill = Biopsy_Occurrence,
                   color = Biopsy_Occurrence)) +
    geom_histogram(alpha = 0.4, position = "identity", bins = 50,
                   aes(y = after_stat(density))) +
    geom_density(alpha = 0.2, linewidth = 1.2) +
    scale_fill_manual(values  = occurrence_colours) +
    scale_color_manual(values = occurrence_colours_dark) +
    labs(
      title    = paste("Distribution of |Δ%| for", neighbor_cell, "around", focal_cell),
      subtitle = "Primary vs Relapse cores (with KDE)",
      x        = "Δ% (|Local proportion − Global proportion|)",
      y        = "Density",
      fill     = "Biopsy Occurrence",
      color    = "Biopsy Occurrence"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12), legend.position = "top")
  print(p2)

  # --- p3: Signed Δ% (zoomed to ±0.02 to focus on the main distribution) ---
  p3 <- ggplot(plot_data_specific,
               aes(x = delta_percent, fill = Biopsy_Occurrence,
                   color = Biopsy_Occurrence)) +
    geom_histogram(alpha = 0.4, position = "identity", bins = 100,
                   aes(y = after_stat(density))) +
    geom_density(alpha = 0.2, linewidth = 1.2) +
    scale_fill_manual(values  = occurrence_colours) +
    scale_color_manual(values = occurrence_colours_dark) +
    coord_cartesian(xlim = c(-0.02, 0.02)) +
    labs(
      title    = paste("Distribution of signed Δ% for", neighbor_cell, "around", focal_cell),
      subtitle = "Primary vs Relapse cores (with KDE)",
      x        = "Δ% (Local proportion − Global proportion)",
      y        = "Density",
      fill     = "Biopsy Occurrence",
      color    = "Biopsy Occurrence"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12), legend.position = "top")
  print(p3)

  # Print per-group summary statistics
  cat("\nSummary statistics for", focal_cell, "→", neighbor_cell, ":\n")
  plot_data_specific %>%
    group_by(Biopsy_Occurrence) %>%
    summarise(
      n_cores                  = n(),
      mean_proportion          = mean(mean_neighbor_proportion, na.rm = TRUE),
      median_proportion        = median(mean_neighbor_proportion, na.rm = TRUE),
      mean_delta_percent_abs   = mean(delta_percent_abs,  na.rm = TRUE),
      median_delta_percent_abs = median(delta_percent_abs, na.rm = TRUE),
      mean_delta_percent       = mean(delta_percent,       na.rm = TRUE),
      median_delta_percent     = median(delta_percent,     na.rm = TRUE)
    ) %>%
    print()

} else {
  cat("No data found for focal:", focal_cell, "| neighbour:", neighbor_cell, "\n")
  cat("Available focal types:\n");    print(unique(plot_data$focal_cell_type))
  cat("Available neighbour types:\n"); print(unique(plot_data$neighbor_cell_type))
}
```

### Save Plots — Primary vs Relapse

```{r save-plots-occurrence}
out_dir <- "/Users/victoireringler/Documents/Singapore/Thesis/cell_distance_analysis/Final_figures/"

ggsave(file.path(out_dir, "Histogram_mean_neighbor_proportion_Treg.pdf"),
       plot = p1, device = "pdf", width = 8, height = 6, units = "in", dpi = 300)

ggsave(file.path(out_dir, "Histogram_delta_percent_abs_Treg.pdf"),
       plot = p2, device = "pdf", width = 8, height = 6, units = "in", dpi = 300)

ggsave(file.path(out_dir, "Histogram_delta_percent_signed_Treg.pdf"),
       plot = p3, device = "pdf", width = 8, height = 6, units = "in", dpi = 300)

cat("Primary vs Relapse plots saved.\n")
```

---

## 6B. Dispersed vs Clustered Analysis (Primary Biopsies Only)

Cores are stratified into tertiles of the **clustering index**:
- **Dispersed** — lowest tertile (most spatially dispersed tumour cells)
- **IQR** — middle tertile (excluded from comparisons)
- **Clustered** — highest tertile (most spatially clustered tumour cells)

### Load & Filter Metadata

```{r load-metadata-cluster}
metadata <- read_xlsx("/Users/victoireringler/Documents/Singapore/Thesis/BCC_mIHC_analysis/survival_ci_per_core.xlsx")

# Same treatment filter as above; restrict to Primary biopsies only
metadata_filt <- metadata %>%
  filter(
    Path              == "DLBC",
    Biopsy_Occurrence == "Primary",
    `Front line treatment (regimen)` %in% treatments_kept,
    clustering_index  <= 50
  )

# Assign tertile-based cluster group labels
vTert <- quantile(metadata_filt$clustering_index, c(0:3/3))
metadata_filt$cluster_group <- with(
  metadata_filt,
  cut(clustering_index, vTert, include.lowest = TRUE,
      labels = c("Dispersed", "IQR", "Clustered"))
)

core_to_cluster <- metadata_filt %>%
  select(Image_ID, cluster_group, clustering_index)

print("Cluster group distribution:")
print(table(metadata_filt$cluster_group))
```

### Merge Metadata

```{r merge-metadata-cluster}
all_mean_proportions_with_meta <- all_mean_proportions %>%
  inner_join(core_to_cluster, by = "Image_ID") %>%
  filter(!is.na(cluster_group))

cat("\nCores kept:", length(unique(all_mean_proportions_with_meta$Image_ID)),
    "out of", length(unique(all_mean_proportions$Image_ID)), "total cores\n")

excluded_cores <- setdiff(unique(all_mean_proportions$Image_ID),
                          unique(all_mean_proportions_with_meta$Image_ID))
if (length(excluded_cores) > 0) {
  cat("Excluded cores (not in metadata_filt):", length(excluded_cores), "\n")
}
```

### Summary Statistics

```{r summary-cluster}
# Overall summary (pooled)
summary_proportions_overall <- all_mean_proportions_with_meta %>%
  group_by(focal_cell_type, neighbor_cell_type) %>%
  summarise(
    mean_neighbor_prop_across_cores = mean(mean_neighbor_proportion, na.rm = TRUE),
    sd_neighbor_prop                = sd(mean_neighbor_proportion,   na.rm = TRUE),
    mean_overall_prop_across_cores  = mean(overall_proportion,       na.rm = TRUE),
    mean_delta_percent_abs          = mean(delta_percent_abs,        na.rm = TRUE),
    mean_delta_percent              = mean(delta_percent,            na.rm = TRUE),
    sd_delta_percent                = sd(delta_percent,              na.rm = TRUE),
    sd_delta_percent_abs            = sd(delta_percent_abs,          na.rm = TRUE),
    n_cores                         = n(),
    .groups = "drop"
  )

# Stratified summary: Dispersed vs Clustered (IQR excluded)
summary_proportions_by_cluster <- all_mean_proportions_with_meta %>%
  filter(cluster_group %in% c("Dispersed", "Clustered")) %>%
  group_by(focal_cell_type, neighbor_cell_type, cluster_group) %>%
  summarise(
    mean_neighbor_prop     = mean(mean_neighbor_proportion, na.rm = TRUE),
    sd_neighbor_prop       = sd(mean_neighbor_proportion,   na.rm = TRUE),
    mean_overall_prop      = mean(overall_proportion,       na.rm = TRUE),
    mean_delta_percent     = mean(delta_percent,            na.rm = TRUE),
    mean_delta_percent_abs = mean(delta_percent_abs,        na.rm = TRUE),
    sd_delta_percent       = sd(delta_percent,              na.rm = TRUE),
    sd_delta_percent_abs   = sd(delta_percent_abs,          na.rm = TRUE),
    n_cores                = n(),
    .groups = "drop"
  )

print(summary_proportions_overall)
print(summary_proportions_by_cluster)
```

### Histogram Plots — Dispersed vs Clustered

```{r plot-cluster}
# ============================================================
# VISUALISE NEIGHBOURHOOD DISTRIBUTIONS: DISPERSED vs CLUSTERED
# ============================================================
plot_data <- all_mean_proportions_with_meta %>%
  filter(cluster_group %in% c("Dispersed", "Clustered"))

focal_cell    <- "Malignant_B_M26"  # Focal cell type (adjust as needed)
neighbor_cell <- "T cell"           # Neighbour cell type (adjust as needed)

plot_data_specific <- plot_data %>%
  filter(focal_cell_type   == focal_cell,
         neighbor_cell_type == neighbor_cell)

cluster_colours      <- c("Dispersed" = "#a6224d", "Clustered" = "#82b4da")
cluster_colours_dark <- c("Dispersed" = "#7f1a3b", "Clustered" = "#4a6f99")

if (nrow(plot_data_specific) > 0) {

  cat("Number of cores per group:\n")
  print(plot_data_specific %>% group_by(cluster_group) %>% summarise(n_cores = n()))

  # --- p1: Mean neighbour proportion ---
  p1 <- ggplot(plot_data_specific,
               aes(x = mean_neighbor_proportion, fill = cluster_group,
                   color = cluster_group)) +
    geom_histogram(alpha = 0.4, position = "identity", bins = 50,
                   aes(y = after_stat(density))) +
    geom_density(alpha = 0.2, linewidth = 1.2) +
    scale_fill_manual(values  = cluster_colours) +
    scale_color_manual(values = cluster_colours_dark) +
    labs(
      title    = paste("Distribution of", neighbor_cell, "proportions around", focal_cell),
      subtitle = "Dispersed vs Clustered cores (with KDE)",
      x        = "Mean neighbour proportion",
      y        = "Density",
      fill     = "Cluster group",
      color    = "Cluster group"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12), legend.position = "top")
  print(p1)

  # --- p2: Absolute Δ% ---
  p2 <- ggplot(plot_data_specific,
               aes(x = delta_percent_abs, fill = cluster_group,
                   color = cluster_group)) +
    geom_histogram(alpha = 0.4, position = "identity", bins = 50,
                   aes(y = after_stat(density))) +
    geom_density(alpha = 0.2, linewidth = 1.2) +
    scale_fill_manual(values  = cluster_colours) +
    scale_color_manual(values = cluster_colours_dark) +
    labs(
      title    = paste("Distribution of |Δ%| for", neighbor_cell, "around", focal_cell),
      subtitle = "Dispersed vs Clustered cores (with KDE)",
      x        = "Δ% (|Local proportion − Global proportion|)",
      y        = "Density",
      fill     = "Cluster group",
      color    = "Cluster group"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12), legend.position = "top")
  print(p2)

  # --- p3: Signed Δ% (zoomed to ±0.02) ---
  p3 <- ggplot(plot_data_specific,
               aes(x = delta_percent, fill = cluster_group,
                   color = cluster_group)) +
    geom_histogram(alpha = 0.4, position = "identity", bins = 100,
                   aes(y = after_stat(density))) +
    geom_density(alpha = 0.2, linewidth = 1.2) +
    scale_fill_manual(values  = cluster_colours) +
    scale_color_manual(values = cluster_colours_dark) +
    coord_cartesian(xlim = c(-0.02, 0.02)) +
    labs(
      title    = paste("Distribution of signed Δ% for", neighbor_cell, "around", focal_cell),
      subtitle = "Dispersed vs Clustered cores (with KDE)",
      x        = "Δ% (Local proportion − Global proportion)",
      y        = "Density",
      fill     = "Cluster group",
      color    = "Cluster group"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12), legend.position = "top")
  print(p3)

  # Print per-group summary statistics
  cat("\nSummary statistics for", focal_cell, "→", neighbor_cell, ":\n")
  plot_data_specific %>%
    group_by(cluster_group) %>%
    summarise(
      n_cores                  = n(),
      mean_proportion          = mean(mean_neighbor_proportion, na.rm = TRUE),
      median_proportion        = median(mean_neighbor_proportion, na.rm = TRUE),
      mean_delta_percent_abs   = mean(delta_percent_abs,  na.rm = TRUE),
      median_delta_percent_abs = median(delta_percent_abs, na.rm = TRUE),
      mean_delta_percent       = mean(delta_percent,       na.rm = TRUE),
      median_delta_percent     = median(delta_percent,     na.rm = TRUE)
    ) %>%
    print()

} else {
  cat("No data found for focal:", focal_cell, "| neighbour:", neighbor_cell, "\n")
  cat("Available focal types:\n");     print(unique(plot_data$focal_cell_type))
  cat("Available neighbour types:\n"); print(unique(plot_data$neighbor_cell_type))
}
```

### Save Plots — Dispersed vs Clustered

```{r save-plots-cluster}
out_dir <- "/Users/victoireringler/Documents/Singapore/Thesis/cell_distance_analysis/Final_figures/"

ggsave(file.path(out_dir, "Histogram_mean_neighbor_proportion_primary_Tcell.pdf"),
       plot = p1, device = "pdf", width = 8, height = 6, units = "in", dpi = 300)

ggsave(file.path(out_dir, "Histogram_delta_percent_abs_primary_Tcell.pdf"),
       plot = p2, device = "pdf", width = 8, height = 6, units = "in", dpi = 300)

ggsave(file.path(out_dir, "Histogram_delta_percent_signed_primary_Tcell.pdf"),
       plot = p3, device = "pdf", width = 8, height = 6, units = "in", dpi = 300)

cat("Dispersed vs Clustered plots saved.\n")
```
