---
title: "Geyer Process Fitting for Spatial Point Patterns"
author: "Shruti"
date: "30/01/2023"
output: html_document
---

## Overview

This script fits a **Geyer saturation process** to spatial point pattern data
for the cell type `Mp2p6n` across multiple tissue samples. For each sample, it:

1. Loads a marked point pattern (`.rds` file)
2. Subsets to the `Mp2p6n` cell type
3. Estimates the optimal interaction radius via the **L-function**
4. Fits a **Geyer saturation model** using profile pseudolikelihood
5. Extracts and saves key model parameters to a CSV

---

## Setup

```{r setup, message=FALSE, warning=FALSE}
# ============================================================
# DEPENDENCIES
# ============================================================
library(spatstat)  # Spatial statistics and point process modelling
```

---

## Load Files

```{r load-files}
# ============================================================
# FILE LOADING
# ============================================================
# Read all .rds file paths from the target directory into a character vector.
# full.names = TRUE returns absolute paths, making downstream reading portable.
files <- list.files(
  path       = "~/Downloads/BCA_RDS_new_total/",
  full.names = TRUE
)

cat("Total files found:", length(files), "\n")
```

---

## Fit Geyer Model

```{r fit-geyer-model, warning=FALSE}
# ============================================================
# MAIN LOOP: Fit Geyer Saturation Process per file
# ============================================================
# Output data frame — columns are named at the end of the loop.
# Pre-allocating with the expected column count (13) avoids repeated reallocation.
df <- data.frame(matrix(ncol = 13, nrow = 0))

for (file_index in seq_along(files)) {

  # ----------------------------------------------------------
  # 1. Load data
  # ----------------------------------------------------------
  file.name     <- files[file_index]
  name.of.file  <- basename(file.name)   # Keep only filename for logging
  MyObjectsData <- readRDS(file.name)    # Each RDS is a marked ppp object

  # ----------------------------------------------------------
  # 2. Subset to cell type of interest: "Mp2p6n"
  # ----------------------------------------------------------
  # Convert marks to factor (required by spatstat subset operations)
  MyObjectsData$marks <- as.factor(MyObjectsData$marks)

  # Retain only points belonging to the "Mp2p6n" mark class
  p        <- subset(MyObjectsData, marks == "Mp2p6n", drop = TRUE)
  p$marks  <- as.factor(p$marks)

  print(name.of.file)

  # ----------------------------------------------------------
  # 3. Skip sparse patterns
  # ----------------------------------------------------------
  # Reliable estimation of second-order statistics and model fitting
  # requires a minimum number of points. Files with fewer than 20
  # points are skipped to avoid degenerate fits.
  if (npoints(p) < 20) {
    cat("  Skipping (fewer than 20 points):", name.of.file, "\n")
    next
  }

  # ----------------------------------------------------------
  # 4. Find optimal interaction radius via the L-function
  # ----------------------------------------------------------
  # The L-function (Besag's transformation of Ripley's K) detects clustering
  # or regularity. The radius r that maximises |L(r) - r| indicates where
  # spatial interaction is strongest — a data-driven choice for interaction range.

  discrep <- function(r) {
    abs(as.function(Lest(p))(r) - r)
  }

  res <- optimise(
    discrep,
    interval = c(0, 50),  # Search range in microns
    maximum  = TRUE        # We want the maximiser, not the minimiser
  )

  cat("  Optimal L-function radius:", round(res$maximum, 2),
      " | Discrepancy:", round(res$objective, 4), "\n")

  # ----------------------------------------------------------
  # 5. Profile pseudolikelihood over a grid of Geyer parameters
  # ----------------------------------------------------------
  # The Geyer saturation process extends the Strauss process by capping
  # the pairwise interaction at a saturation count `sat`, allowing it to
  # model both hard-core repulsion and clustered patterns.
  #
  # Parameter grid:
  #   r   — interaction radius (1–30 µm in 0.5 µm steps)
  #   sat — saturation threshold (1, 2, or 3 neighbours)
  #
  # profilepl() selects the (r, sat) pair that maximises the
  # penalised pseudologlikelihood (aic = FALSE → use PL, not AIC).

  rvals      <- seq(1, 30, by = 0.5)
  param.grid <- expand.grid(
    r   = rvals,
    sat = seq(1, 3, by = 1)
  )

  fitp <- profilepl(param.grid, Geyer, p, aic = FALSE)

  # ----------------------------------------------------------
  # 6. Extract best-fitting model and key parameters
  # ----------------------------------------------------------
  X <- as.ppm(fitp)  # Best-fitting ppm object

  # The Geyer model coefficient for log(gamma) is the second element.
  # Exponentiating gives the interaction parameter gamma:
  #   gamma > 1  -> clustering  (attraction)
  #   gamma < 1  -> regularity  (inhibition)
  #   gamma = 1  -> CSR         (no interaction)
  gamma <- exp(X$coef[2])

  # Retrieve the selected interaction radius and saturation parameter
  interaction.distance <- fitp$fit$interaction$par[1]  # r   (microns)
  saturation.parameter <- fitp$fit$interaction$par[2]  # sat

  # ----------------------------------------------------------
  # 7. Collect results for this file
  # ----------------------------------------------------------
  # Coefficients 1–7 cover the second-order polynomial trend surface:
  # Intercept, x, y, x², x·y, y², log(Interaction)
  fit.row <- c(
    name.of.file,
    as.numeric(X$coef[1:7]),           # Trend + interaction coefficients
    as.numeric(gamma),                 # Interaction strength
    as.numeric(res$maximum),           # Optimal L-function radius
    as.numeric(res$objective),         # Max L-function discrepancy
    as.numeric(interaction.distance),  # Geyer r
    as.numeric(saturation.parameter)   # Geyer sat
  )

  df <- rbind(df, as.data.frame(t(fit.row), stringsAsFactors = FALSE))
}

# ============================================================
# COLUMN NAMING
# ============================================================
names(df) <- c(
  "Name",
  "Intercept", "x", "y", "I(x^2)", "I(x*y)", "I(y^2)",
  "log(Interaction)",
  "gamma",
  "Opt_L_radius",
  "L_discrepancy",
  "Interaction_Distance",
  "Saturation_Parameter"
)
```

---

## Results

```{r display-results}
# ============================================================
# DISPLAY & EXPORT
# ============================================================
print(df)

# Save results to CSV (row.names = FALSE avoids a redundant integer index column)
write.csv(
  df,
  file      = "~/OneDrive - National University of Singapore/PhD_code/Spatial project/Data/BCA/BCA_Mp2p6n_Geyer_bestPL.csv",
  row.names = FALSE
)

cat("Results written to CSV successfully.\n")
```

---

## Clustering Index Derivation

```{r clustering-index}
# ============================================================
# CLUSTERING INDEX (CI) DERIVATION
# ============================================================
# CI integrates two complementary spatial statistics:
#   - gamma: Geyer interaction parameter (model-based, directional)
#   - L_discrepancy: max |L(r) - r| (empirical, non-parametric)
#
# CI = (gamma + L_discrepancy) / 2
#
# For patients with multiple cores, gamma and L_discrepancy are
# averaged across cores before computing CI, giving a single
# patient-level summary statistic.
#
# Higher CI -> more clustered
# Lower CI  -> more dispersed

# ── Load reference file mapping core names to patient IDs ──
ref <- read.csv("BCA_Mp2p6n_Geyer_with_patients.csv")

# ── Convert numeric columns ──
df$gamma        <- as.numeric(df$gamma)
df$L_discrepancy <- as.numeric(df$L_discrepancy)

# ── Join patient IDs ──
df <- df %>%
  left_join(ref %>% select(Name, Patient), by = "Name")

cat("Cores matched to patient ID:", sum(!is.na(df$Patient)),
    "of", nrow(df), "\n")

# ── Compute patient-level CI by averaging across cores ──
# CI(av) = [L(1) + L(2) + ... + L(n) + gamma(1) + gamma(2) + ... + gamma(n)] / 2n
df_ci <- df %>%
  filter(!is.na(Patient)) %>%
  group_by(Patient) %>%
  summarise(
    avg_gamma        = mean(gamma,         na.rm = TRUE),
    avg_L_discrepancy = mean(L_discrepancy, na.rm = TRUE),
    n_cores          = n(),
    .groups          = "drop"
  ) %>%
  mutate(CI = (avg_gamma + avg_L_discrepancy) / 2)

# ── Export ──
write.csv(
  df_ci,
  file      = "BCA_Mp2p6n_CI_per_patient.csv",
  row.names = FALSE
)

cat("CI derivation complete.\n")
print(summary(df_ci$CI))
```
